# phase_mnv_rs

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = "", collapse = TRUE)

run_cmd <- function(cmd, args = character()) {
  if (!file.exists(cmd) && !nzchar(Sys.which(cmd))) {
    return(sprintf(
      "%s not found. Run `make release` before rendering README.md.",
      cmd
    ))
  }
  out <- suppressWarnings(system2(cmd, args, stdout = TRUE, stderr = TRUE))
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    out <- c(out, sprintf("[exit status %s]", status))
  }
  paste(out, collapse = "\n")
}

run_tool <- function(cmd, args = "--help") {
  run_cmd(cmd, args)
}

phase_mnv_rs <- file.path("target", "release", "phase_mnv_rs")
phase_compare <- file.path("target", "release", "phase_compare")
unphase_vcf <- file.path("target", "release", "unphase_vcf")
fermi_lite_assemble <- file.path("target", "release", "fermi_lite_assemble")
bam_error_model <- file.path("target", "release", "bam_error_model")
fixture_ref <- file.path("tests", "fixtures", "ref.fa")

fence <- function(x) {
  cat("```text\n")
  cat(x)
  cat("\n```\n")
}

external_example <- function() {
  if (!identical(Sys.getenv("PHASE_MNV_RUN_EXTERNAL"), "1")) {
    cat(paste(
      "Set `PHASE_MNV_EXAMPLE_VCF`, `PHASE_MNV_EXAMPLE_REF`, and",
      "`PHASE_MNV_EXAMPLE_SAMPLE`, then run `make readme-external-example`",
      "to run a local larger VCF through the Rust binary.",
      "",
      "This is intentionally disabled in the committed README so no local/private",
      "paths or data names are embedded.",
      sep = "\n"
    ))
    return(invisible(FALSE))
  }

  vcf <- Sys.getenv("PHASE_MNV_EXAMPLE_VCF")
  ref <- Sys.getenv("PHASE_MNV_EXAMPLE_REF")
  sample <- Sys.getenv("PHASE_MNV_EXAMPLE_SAMPLE")
  missing <- names(which(!nzchar(c(
    PHASE_MNV_EXAMPLE_VCF = vcf,
    PHASE_MNV_EXAMPLE_REF = ref,
    PHASE_MNV_EXAMPLE_SAMPLE = sample
  ))))
  if (length(missing)) {
    stop("Missing required external README example variable(s): ", paste(missing, collapse = ", "))
  }
  if (!file.exists(vcf)) stop("External README example VCF does not exist")
  if (!file.exists(ref)) stop("External README example reference FASTA does not exist")

  tmp <- tempfile("phase-mnv-readme-")
  dir.create(tmp)
  on.exit(unlink(tmp, recursive = TRUE), add = TRUE)
  rs_out <- file.path(tmp, "rust.vcf")

  rs <- suppressWarnings(system2(
    phase_mnv_rs,
    c("-q", "-r", ref, "-s", sample, "-o", rs_out, vcf),
    stdout = TRUE,
    stderr = TRUE
  ))
  if (!is.null(attr(rs, "status")) && attr(rs, "status") != 0) {
    stop("Rust external README example failed: ", paste(rs, collapse = "\n"))
  }

  lines <- readLines(rs_out, warn = FALSE)
  n_records <- sum(!startsWith(lines, "#"))
  fence(paste(
    "external VCF Rust example passed",
    sprintf("emitted_records=%d", n_records),
    sep = "\n"
  ))
  invisible(TRUE)
}
```

`phase_mnv_rs` is a phased VCF/BCF haplotype merger. It takes variants carried
on the same phased haplotype/phase set and emits normalized `TYPE=MNV` or
`TYPE=COMPLEX` records.

## Scope at a glance

| Area | Status |
| --- | --- |
| Default output | Derived MNV/COMPLEX records only (`--emit mnv`). |
| Canonical implementation | Rust (`src/main.rs`), built on `rust-htslib`. |
| Native dependencies | Vendored C libraries only when useful, exposed through generated/bindgen-backed FFI wrappers. |
| Extra Rust modes | BCF/BGZF output, all-sites output, experimental BAM/CRAM phasing, codon-aware recomposition, native phase comparison, native VCF/BCF unphasing, empirical BAM error-model summaries, and experimental fermi-lite local assembly bindings. |

The Rust binary uses `rust-htslib` for VCF/BCF/BAM/CRAM APIs and only drops to
`rust_htslib::htslib` where exact htslib behavior is needed. Rust-only features
are documented as experimental unless explicitly validated against an upstream
reference.

## Quick start

```bash
. "$HOME/.cargo/env"
make release

target/release/phase_mnv_rs \
  --reference ref.fa \
  --sample S1 \
  --output mnv.vcf \
  input.vcf.gz
```

Build or install variants:

```bash
make install          # install Rust CLIs to ~/.local/bin
make static-release   # static Linux Rust binary when supported
make install-static   # install static Linux binary when supported
# cargo also builds phase_compare, unphase_vcf, fermi_lite_assemble, and bam_error_model
```

Override the Rust target when needed:

```bash
make release TARGET=aarch64-apple-darwin
make static-release STATIC_TARGET=x86_64-unknown-linux-gnu
```

## Modes and output

| Option | Meaning |
| --- | --- |
| `--emit mnv` | Default. Emit only derived `TYPE=MNV` / `TYPE=COMPLEX` records. |
| `--emit all-sites` | Preserve all input records/header and update `GT:PS` where phasing is available. Does not append MNV/COMPLEX records yet. |
| `--mnv-algorithm proximity` | Default window/phase-set recomposition. |
| `--mnv-algorithm nirvana-codon --codon-map codons.tsv` | Initial SNV-only same-codon recomposition slice; not full Nirvana parity. |
| `--phase-from-bam reads.bam` | Experimental indexed BAM/CRAM-backed phasing before output. |

The codon map is BED-like: `CHROM START0 END0 TRANSCRIPT CODON_ID`. In
`nirvana-codon` mode, phased SNV observations are recomposed only when at least
two observations on the same haplotype/phase set share a transcript/codon key.

Output format is inferred from `-o/--output`:

```text
-o out.vcf      plain VCF
-o out.vcf.gz   BGZF-compressed VCF
-o out.vcf.bgz  BGZF-compressed VCF
-o out.bcf      BCF
```

Use `--threads N` or `-@ N` for htslib/BGZF worker threads on compressed input
and output. Stdout remains plain VCF.

## Citation

Please cite the upstream methods that match the parts of `phase_mnv_rs` you use.

For normalized `REF`/`ALT` representation:

> Tan A, Abecasis GR, Kang HM. **Unified representation of genetic variants.**
> *Bioinformatics.* 2015;31(13):2202-2204.
> [doi:10.1093/bioinformatics/btv112](https://doi.org/10.1093/bioinformatics/btv112).

For workflows that run external WhatsHap via
`scripts/phase_from_bam_then_mnv.sh`:

> Martin M, Patterson M, Garg S, Fischer SO, Pisanti N, Klau GW, Schoenhuth A,
> Marschall T. **WhatsHap: fast and accurate read-based phasing.**
> *bioRxiv.* 085050.
> [doi:10.1101/085050](https://doi.org/10.1101/085050).

For the read-backed weighted haplotype assembly/MEC ideas behind the
experimental Rust `--phase-from-bam --phase-algorithm mec` mode:

> Patterson M, Marschall T, Pisanti N, van Iersel L, Stougie L, Klau GW,
> Schönhuth A. **WhatsHap: Weighted Haplotype Assembly for Future-Generation
> Sequencing Reads.** *Journal of Computational Biology.* 2015;22(6):498-509.
> [doi:10.1089/cmb.2014.0157](https://doi.org/10.1089/cmb.2014.0157).

The Rust BAM phaser is WhatsHap-inspired, not a full WhatsHap/PedMEC clone. Use
WhatsHap itself when you need established WhatsHap behavior and cite the
appropriate WhatsHap publication for that workflow.

## CLI help

The help text below is generated directly from the built binaries when
`README.md` is rendered from `README.Rmd`.

### Rust binary (`phase_mnv_rs`)

```{r rust-help, results='asis'}
fence(run_tool(file.path("target", "release", "phase_mnv_rs")))
```

### Phase comparison binary (`phase_compare`)

```{r phase-compare-help, results='asis'}
fence(run_tool(phase_compare))
```

### Native unphasing helper (`unphase_vcf`)

```{r unphase-help, results='asis'}
fence(run_tool(unphase_vcf))
```

### fermi-lite smoke/utility binary (`fermi_lite_assemble`)

```{r fermi-lite-help, results='asis'}
fence(run_tool(fermi_lite_assemble))
```

### BAM empirical error model helper (`bam_error_model`)

```{r bam-error-model-help, results='asis'}
fence(run_tool(bam_error_model))
```

## Examples

These examples are executed when `README.md` is rendered from `README.Rmd`, so
shown output is produced by the current binary instead of copied by hand.

### Adjacent phased SNVs become `TYPE=MNV`

Command:

```bash
target/release/phase_mnv_rs -r tests/fixtures/ref.fa --no-header -q tests/fixtures/phased_mnv.vcf
```

Output:

```{r example-mnv, results='asis'}
fence(run_tool(
  phase_mnv_rs,
  c("-r", fixture_ref, "--no-header", "-q", file.path("tests", "fixtures", "phased_mnv.vcf"))
))
```

### Mixed SNV/indel blocks become `TYPE=COMPLEX`

Command:

```bash
target/release/phase_mnv_rs -r tests/fixtures/ref.fa --no-header -q tests/fixtures/complex.vcf
```

Output:

```{r example-complex, results='asis'}
fence(run_tool(
  phase_mnv_rs,
  c("-r", fixture_ref, "--no-header", "-q", file.path("tests", "fixtures", "complex.vcf"))
))
```

### Experimental Rust BAM-backed phasing before MNV construction

The Rust binary can now phase from an indexed BAM/CRAM before constructing MNVs:

```bash
target/release/phase_mnv_rs \
  --reference ref.fa \
  --sample S1 \
  --phase-from-bam reads.bam \
  --max-gap 100 \
  --output mnv.vcf \
  chromosome.vcf.gz
```

This Rust-only mode ignores input `GT` phase and `FORMAT/PS`, extracts allele
observations from reads with `rust-htslib`, performs deterministic read selection
(default `--phase-max-coverage 15`), builds connected phase components, and by
default solves the exact weighted diploid single-sample MEC objective per
component (`--phase-algorithm mec`). The earlier pairwise parity heuristic is
still available as `--phase-algorithm greedy`. It is intentionally described as
experimental: it is WhatsHap-inspired, but not yet a full clone of WhatsHap's
PedMEC implementation or all WhatsHap options.

### BAM-backed phasing with external WhatsHap before MNV construction

For the established upstream phaser, or for comparison against the Rust phaser,
use the helper workflow below. It first converts all GT separators from `|` to
`/`, drops `FORMAT/PS` and `FORMAT/PQ` by default, runs `whatshap phase`, then
runs `phase_mnv_rs` on the WhatsHap-phased VCF.

Install the external phasing/indexing tools if needed:

```bash
micromamba create -y -n phase-mnv-whatshap -c conda-forge -c bioconda \
  whatshap bcftools samtools
micromamba run -n phase-mnv-whatshap whatshap --version
```

Run the local BAM-backed phasing workflow with explicit local paths:

```bash
./scripts/phase_from_bam_then_mnv.sh \
  --reference ref.fa \
  --bam reads.bam \
  --vcf chromosome.vcf.gz \
  --sample S1 \
  --max-gap 100 \
  --out-dir local_runs/S1-chromosome
```

The helper writes:

```text
PREFIX.unphased.vcf.gz
PREFIX.whatshap.vcf.gz
PREFIX.phase_mnv.vcf
PREFIX.whatshap.log
PREFIX.phase_mnv.log
```

The unphasing step can also be run directly:

```bash
target/release/unphase_vcf input.vcf.gz | bgzip -c > input.unphased.vcf.gz
bcftools index -f input.unphased.vcf.gz
```

### Optional local larger VCF example

The committed README only runs tracked fixtures. To exercise a larger local VCF
while rendering this README, provide explicit paths via environment variables;
the rendered external-example output intentionally omits those paths.

```bash
PHASE_MNV_EXAMPLE_VCF=input.vcf.gz \
PHASE_MNV_EXAMPLE_REF=ref.fa \
PHASE_MNV_EXAMPLE_SAMPLE=S1 \
make readme-external-example
```

```{r example-external, results='asis'}
external_example()
```

## Test

CI and default tests use only tracked fixtures under `tests/fixtures/`.

```bash
make test                  # Rust behavior, unphase_vcf, output-format, bcftools norm, BAM phasing, phase_compare, fermi-lite FFI, BAM error model, and negative tests
make compare-whatshap-phase # optional WhatsHap truth path compared with native phase_compare
```

For private or larger local datasets, pass paths from your shell; no private
paths are embedded in the repository or rendered README.

## README generation

`README.md` is generated from `README.Rmd` so the CLI help and fixture example
outputs stay synchronized with the installed tools:

```bash
make readme
```

This requires R with the `knitr` package. The `readme` target builds the Rust
CLIs first, then renders `README.Rmd`. For a local larger VCF example, set
`PHASE_MNV_EXAMPLE_VCF`, `PHASE_MNV_EXAMPLE_REF`, and
`PHASE_MNV_EXAMPLE_SAMPLE`, then run:

```bash
make readme-external-example
```

## Reference implementations

Reference implementations for comparison/compatibility work are not vendored in
this repository. Clone them when needed. The fermi-lite library is vendored
separately under `vendor/fermi-lite` as an optional local assembly substrate,
not as a phase/MNV reference implementation:

```bash
./scripts/clone_reference_impls.sh
```

### Optional WhatsHap phase-concordance comparison

The external phasing comparison uses WhatsHap plus the native `phase_compare`
binary. It builds a truth path by unphasing the tiny tracked VCF and running
external `whatshap phase`. It builds the query path by running Rust
`phase_mnv_rs --emit all-sites --phase-from-bam` directly on the same VCF/BAM.
Then `phase_compare` reports common heterozygous sites, intersection PS blocks,
assessed adjacent pairs, phase matches, switch errors, and blockwise Hamming
rate:

```bash
make compare-whatshap-phase
```

This is intentionally faster and narrower than hap.py: it measures phasing and
PS/block concordance, not generic variant-call equivalence. The comparison
script strips bcftools/producer command headers and path-bearing
`phase_mnv`/reference records from generated VCFs before comparison, so local
filesystem paths are not retained in comparison VCF headers.

The script discovers `whatshap` from `PATH`, then falls back to a micromamba
environment named `phase-mnv-whatshap`. Override with `WHATSHAP_BIN` or
`WHATSHAP_ENV` when needed.

Detailed semantics and validation notes are in:

- [`docs/semantics.md`](docs/semantics.md)
- [`docs/validation.md`](docs/validation.md)

### Experimental fermi-lite local assembly binding

The repository vendors fermi-lite source under `vendor/fermi-lite` and builds it
through `build.rs` for future local read-evidence adjudication. Bindings are
generated from `fml.h` with bindgen when libclang is available, with a checked-in
narrow fallback for portability (`PHASE_MNV_REQUIRE_BINDGEN=1` forces bindgen).
The fermi-lite wrapper is currently enabled on x86_64 targets because the
vendored upstream `ksw.c` path requires SSE2. The current exposed Rust path is
intentionally small: `src/fermi_lite.rs` wraps
`fml_opt_init`, `fml_assemble`, and
`fml_utg_destroy`, and
`fermi_lite_assemble` is a smoke/utility binary for assembling supplied local
read sequences into FASTA unitigs. It can pass FASTQ/base qualities through to
fermi-lite when `--fastq --ec-k 0` or another non-negative `--ec-k` is used.

`bam_error_model` is a separate helper that learns simple empirical mismatch,
insertion, and deletion summaries from a BAM/CRAM versus a reference, without a
MAPQ filter by default. It reports base-quality and MAPQ bins so a future
`phase_adjudicate` path can separate local assembly from quality-aware evidence
scoring. Known variant sites are not masked yet, so mismatch rates from this
helper should be interpreted as sequencing-error-plus-variation unless regions
are restricted to homozygous-reference/high-confidence sites.

Neither helper is yet integrated into `phase_compare` or a final
`phase_adjudicate` workflow.

If you use fermi-lite-backed local assembly results, cite the FermiKit paper
recommended by fermi-lite (Li 2015, Bioinformatics; doi:10.1093/bioinformatics/btv440).

The most relevant upstream tools are:

- `whatshap phase` / `whatshap haplotag` — established read-backed phasing
  reference for BAM/CRAM validation and external phasing workflows.
- `phase_compare` — in-repository native phase-concordance comparator for
  WhatsHap-derived all-sites truth versus Rust BAM-backed all-sites output.
- `vt normalize` — normalization reference for left-aligned + parsimonious
  representation.
- `bcftools norm` — useful validator for emitted normalized records.
- `Illumina Nirvana` — codon/transcript-aware SNV-only MNV recomposition
  reference for future benchmarking of annotation-driven recomposition rules.
- `fermi-lite` / FermiKit — optional local reassembly substrate for future
  read-evidence adjudication of difficult phasing disagreements.

## CI

GitHub Actions builds and tests the Rust project on Linux and macOS:

- Rust release/static where supported
- behavior and negative/failure-mode tests for the Rust binaries
- Rust `bcftools norm` validation of emitted normalized records
- Linux WhatsHap-derived all-sites phase comparison with native `phase_compare`
- native `unphase_vcf`, fermi-lite FFI, and BAM error-model smoke coverage
- binary artifact upload with SHA256 sums

## Notes

- Reads VCF/BCF via htslib/rust-htslib.
- Uses phased diploid GT and `FORMAT/PS`.
- Emits `TYPE=MNV` for pure SNV blocks and `TYPE=COMPLEX` for blocks including indels.
- Normalizes internally with the Tan, Abecasis & Kang 2015 left-aligned + parsimonious rules (doi:10.1093/bioinformatics/btv112).
- Does not require a separate `vt normalize` or `bcftools norm` pass for emitted records.
